---
title: "temperature vignette"
author: 
  - name: Zijun Gao
    affiliation: Department of Statistics, Stanford University
  - name: Trevor Hastie
    affiliation: Department of Statistics and Department of Biomedical Data Science, Stanford University
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    toc: no
    toc_depth: 3
    number_sections: yes
    toc_float:
      collapsed: no
    code_folding: show
    theme: cerulean
vignette: >
  %\VignetteIndexEntry{temperature vignette} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r global setting, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = FALSE
)
```

In this vignette, we will introduce how to use the R package _temperature_ to study the distribution of body temperature. We first install and load the package.
```{r}
library("temperature")
```

We prepare a data frame containing features of subjects. The data frame should contain the following five items: 

  * time: the time to compute the temperature at. The \code{time} should be a character of the format hh:mm, e.g., "08:30" and "12:00". The \code{time} should be no earlier than 07:00 or later than 18:00 within a day.
  * gender: gender of the subject. The \code{gender} should be a character. There are three possible values: "male", "female.pre" for ladies before menopause, and "female.post" for ladies after menopause. If the menopause condition in not available, a default rule regards female no younger than 40 as post-menopause.
  * age: age of the subject. The \code{age} should be numeric and will be rounded to the nearest integer. The \code{age} should be no smaller than 20 or larger than 80.
  * height: height of the subject (in meter). The \code{height} should be numeric and will be rounded to two decimal places. The \code{height} should be no smaller than 1.38 or larger than 2.13.
  * weight: weight of the subject (in kilogram). The \code{weight} should be numeric and will be rounded to the nearest integer. The \code{weight} should be no smaller than 30 or larger than 181.

In below, we create an example data frame with three subjects.
```{r data}
data = data.frame(time = c("08:00", "12:00", "14:00"), gender = c("male", "female.pre", "female.post"), age = c(30, 30, 70), height = c(1.8, 1.78, 1.6), weight = c(70, 60, 50.5))
data
```


# Percentile to temperature

In this section, we display how to compute the body temperature corresponding to a target percentile for a given subject. 

We add a column named _percentile_ to the current data frame. Next we use the function _percentile.to.temperature_ in the package to compute the temperature.
```{r percentile to temperature}
data$percentile = c(50,95,99) # 50%, 95%, 99% percentiles for the three subjects, respectively
percentile.to.temperature(data)
```
The predicted $50\%$ percentile body temperature for the first subject is $97.80 ^{\circ}F$. Similar for other subjects at given percentiles.

<!-- Given a subject, we can plot the predicted temperatures from the $1\%$ percentile to the $99\%$ percentile. The figure below demonstrates the temperature curve of the first subject in the data frame. -->
<!-- ```{r percentile to temperature curve} -->
<!-- data.curve = data[rep(1,99),] -->
<!-- data.curve$percentile = seq(1,99) -->
<!-- plot(1:99, percentile.to.temperature(data.curve), pch = 16, xlab = "percentile (%)", ylab = "tempertaure (\u00B0F)") -->
<!-- ``` -->


# Temperature to percentile

In this section, we display how to compute the percentile corresponding to the body temperature of a given subject. 

We add a column named _temperature_ to the current data frame. Next, we use the function _temperature.to.percentile_ in the package to compute the percentile.
```{r temperature to percentile}
data$temperature = c(98,98,98)  # body temperatures of all three subjects are 98F
temperature.to.percentile(data)
```
For the first subject, the body temperature $98 ^{\circ}F$ is around the $67\%$ percentile. Similar for other subjects. We remark that, if the body temperatures fall below the $1\%$ percentile or above the $99\%$ percentile, the function will output $-1$ and $101$, respectively.

# Cumulative distribution of temperature

Given a subject, we can plot the cumulative distribution of his or her temperature. Below we display the cumulative distribution curve of the first subject in the data frame and highlight the observed temperature.

```{r temperature to percentile curve}
data.curve = data[rep(1,99),]
data.curve$percentile = seq(1,99)
plot(percentile.to.temperature(data.curve), 1:99, pch = 16, xlab = "tempertaure (\u00B0F)", ylab = "percentile (%)", main = "cumulative distribution of temperature")
points(x = data$temperature[1], y = temperature.to.percentile(data)[1], col = "red", pch = 16, cex = 1.5)
text(label = "observed temperature", x = data$temperature[1]-0.5, y = temperature.to.percentile(data)[1]+5)
```
